---
title: "power_4"
author: "Elias"
date: "May 21, 2018"
output: html_document
---
# __Powerful Data Project__ Part III: Are we better lifters than our moms and dads?




```{r}
library(lme4)
library(tidyr)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(gtools)
library(tinytex)

open<-read.csv("~/power/data/open/openpowerlifting2.csv",head=TRUE,stringsAsFactors = FALSE) #reading in a re-formatted and merged version of the two csv documents that come from the openpowerlifting database
# open<-read.csv("~/power/data/open/openpowerlifting.csv",head=TRUE,stringsAsFactors = FALSE) #reading in the first csv document
# meets<-read.csv("~/power/data/open/meets.csv",head=TRUE,stringsAsFactors = FALSE) #reading in the one containing meets
# all<-merge(open,meets,by="MeetID") #merging them by column (MeetID, the only repeated column)
# # # 
str(open) #getting the structure of the new document

```

Obviously this is a fantastic and well kept dataset! __Great job__ openpowerlifting.com!

I want to draw your attention that there are so many columns in this dataset, each containing unique information about each lifter. This will give us a lot of power to draw some observations about patterns of skills and potentials of the lifters. First I want to start by breaking them up into age ranges. I don't like the ones drawn by more federations for my purposes. I want to have a bit higher resolution, so I will break them up into 5 year age gaps and see how things go from there.

I am also breaking them up into men and women for physiological reasons. I will be asking questions about max lift strength, so I don't want to obscure the findings for either of those groups.

```{r}
#Creating a vector of ages in increments of 5
# group<-seq(from=range(all$Age,na.rm=TRUE)[[1]],
#             to=range(all$Age,na.rm=TRUE)[[2]],
#             by=5)
# 
# group_range<-c() #creating a vector that will hold those ranges
# group2<-c(group-1,100) #creating a second vector that has 100 as the maximum of it
# for(i in 1:length(group)){ #creating teh group-range vector that will have info about a range, rather than a single digit
#  if(i==length(group)){group_range[i]<-paste(c(group[i],"100"),collapse="-")
#  next()
#  }
#  group_range[i]<-(paste(c(group[i],group2[i+1]),collapse="-"))
# } #making these into ranges rather than just the minimums
# 
# #Run the following only once as it takes a while (20 mins)
# all<-cbind(all,Group=NA) # Adding a column of Group to the big dataset
# all$Age<-round(all$Age) #rounding age because the following code doesn't deal well with .5's and they are stupid anyways
# 
# #adding a column of ranges
# for(i in 1:length(all$Age)){
#  if(is.na(all$Age[i])){next()} #remove this estimation from all NAs
#  for(j in 1:length(group)){
#    if(all$Age[i]>=group[j]&&all$Age[i]<=group2[j+1]){ #iterate over groups and find which group Age this falls into
#    all$Group[i]<-group_range[j] #assign the necessary group range to that age
#    }
#  }
# }
# 
# #Adding weight groups (all freaking federations have different weightclasses so it's easier to actually group myself)
# groupkg<-seq(from=range(all$BodyweightKg,na.rm=TRUE)[[1]],
#             to=range(all$BodyweightKg,na.rm=TRUE)[[2]],
#             by=10)
# groupkg<-round(groupkg)
# 
# groupkg_range<-c() #creating a vector that will hold those ranges
# groupkg2<-c(groupkg-1,260) #creating a second vector that has 260 as the maximum of it
# for(i in 1:length(groupkg)){ #creating teh group-range vector that will have info about a range, rather than a single digit
#  if(i==length(groupkg)){groupkg_range[i]<-paste(c(groupkg[i],"260"),collapse="-")
#  next()
#  }
#  groupkg_range[i]<-(paste(c(groupkg[i],groupkg2[i+1]),collapse="-"))
# } #making these into ranges rather than just the minimums
# 
# #Run the following only once as it takes a while (20 mins)
# all<-cbind(all,GroupKG=NA) # Adding a column of Group to the big dataset
# all$BodyweightKg<-round(all$BodyweightKg) #rounding age because the following code doesn't deal well with .5's and they are stupid anyways
# 
# #adding a column of ranges
# for(i in 1:length(all$BodyweightKg)){
#  if(is.na(all$BodyweightKg[i])){next()} #remove this estimation from all NAs
#  for(j in 1:length(groupkg)){
#    if(all$BodyweightKg[i]>=groupkg[j]&&all$BodyweightKg[i]<=groupkg2[j+1]){ #iterate over groups and find which group BW falls
#    all$GroupKG[i]<-groupkg_range[j] #assign the necessary group range to that age
#    }
#  }
# }
# 
# 
# #  # #writing table for future use because the chunk above takes FOREVER (30 min-ish)
# write.table(all,"~/power/data/open/openpowerlifting2.csv",quote=FALSE,row.names = FALSE,sep=',')


ord<-mixedorder(open$Group) #sorting by the group column so that they plot by age
open2<-open[ord,] #aplying that to the whole dataframe
open2$Group<-factor(open2$Group,levels=unique(open2$Group)) #arranging the factors now.

men<-open2 %>% #making a table of men who have completed full competitions and have Age data
  filter(BestBenchKg>0 & BestDeadliftKg>0 & BestSquatKg>0) %>% 
  filter(!is.na(Age)) %>%
  filter(Sex=="M") %>% 
  select(Equipment,Age,BodyweightKg,BestSquatKg,BestBenchKg,BestDeadliftKg,TotalKg,Wilks,Group,WeightClassKg)

women<-open2 %>% #same for women
  filter(BestBenchKg>0 & BestDeadliftKg>0 & BestSquatKg>0) %>% 
  filter(!is.na(Age)) %>% 
  filter(Sex=="F") %>% 
  select(Equipment,Age,BodyweightKg,BestSquatKg,BestBenchKg,BestDeadliftKg,TotalKg,Wilks,Group,WeightClassKg)

```